Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-06-07 doc/arxiv_version/paper.tex
Cleanup

% KEYWORDS:
% model counting
% first-order logic
% recursion
% knowledge compilation

% TOPICS:
% KR & ML: probabilistic reasoning and learning
% KR & ML: statistical relational learning
% (no) computational aspects of knowledge representation
% (no) knowledge representation languages
% (no) KR & ML: graphical models for knowledge representation and reasoning

@@

In this work, we show how to relax these restrictions in a way that results in
an algorithm capable of handling more instances in a lifted manner. The ideas
presented in this paper are implemented in \textsc{Crane}---an extension of the
(W)FOMC algorithm \textsc{ForcLift}. The differences in how these algorithms
operate are depicted in \cref{fig:process}. Compilation is performed by applying
various \emph{(compilation) rules} to the input (or some derivative) formula,
gradually constructing a circuit (in the case of \textsc{ForcLift}) or a graph
(in the case of \textsc{Crane}). \textsc{ForcLift} applies compilation rules via
greedy search, whereas \textsc{Crane} also supports a hybrid search algorithm
that applies some rules greedily and some using breadth-first
search.\footnote{In the current implementation, the rules applied non-greedily
  are: atom counting, inclusion-exclusion, independent partial groundings,
  Shannon decomposition, shattering, and two new rules described in
  \cref{sec:dr,sec:ref}---see previous work~\cite{DBLP:conf/ijcai/BroeckTMDR11}
  for more information about the rules.} This alternative was introduced because
there is no reason to expect greedy search to be complete. Another difference is
that---in \textsc{Crane}---the product of compilation is not directly evaluated
but transformed into a collection of functions on domain sizes. Hence, our
approach is reminiscent of previous work on lifted inference via compilation to
C++ programs~\cite{DBLP:conf/kr/KazemiP16} and the broader area of functional
synthesis~\cite{DBLP:conf/cav/GoliaRM20,DBLP:conf/pldi/KuncakMPS10,sanathanan1963transfer}.

@@

Our representation of FOMC instances builds on the format used internally by
\textsc{ForcLift}, some aspects of which are described by Van den Broeck et
al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate
sentences in a variant of function-free many-sorted first-order logic with
equality to this internal format. We use lowercase Latin letters for predicates
(e.g., $\predicate$) and constants (e.g., $x$), uppercase Latin letters for
variables (e.g., $X$), and uppercase Greek letters for domains (e.g., $\Delta$).
Sometimes we write predicate $\predicate$ as $\predicate/n$, where
$n \in \mathbb{N}^{+}$ is the \emph{arity} of $\predicate$. An \emph{atom} is
$\predicate(x_1, \dots, x_n)$ for some predicate $\predicate/n$ and terms
$x_{1}, \dots, x_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation thereof (denoted by
$\neg \predicate(x_1, \dots, x_n)$). Let $\mathcal{D}$ be the set of all
relevant (finite) domains. Initially, $\mathcal{D}$ contains all domains
mentioned by the input formula. During compilation, new domains are added to
$\mathcal{D}$. Each new domain is interpreted as a subset of another domain in
$\mathcal{D}$.

@@

The model count of a formula $\phi$ also depends on the sizes of the domains in
$\phi$. Let $\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain
  size function} that maps each domain to a non-negative integer.

\begin{example}
  Let $\phi$ be as in \cref{example:first} and $\Gamma = \Delta = \{\,1, 2\,\}$,
  i.e., $\sigma(\Gamma) = \sigma(\Delta) = 2$. There are $2^{2 \times 2} = 16$
  possible relations between $\Gamma$ and $\Delta$. Let us count how many of
  them satisfy the conditions imposed on predicate $\predicate$. The empty
  relation does. All four relations of cardinality one (e.g.,
  $\{\, (1, 1) \,\}$) do too. Finally, there are two relations of cardinality
  two---$\{\, (1, 1), (2, 2) \,\}$ and $\{\, (1, 2), (2, 1) \,\}$---that satisfy
  the conditions as well. Therefore, the FOMC of $(\phi, \sigma)$ is 7.
  Incidentally, the FOMC of $\phi$ counts partial injections from $\Gamma$ to
  $\Delta$. We will continue to use the problem of counting partial injections
  as the main running example.
\end{example}

@@

% Original domain recursion
Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} imposes stringent conditions
on the input formula to ensure that the expanded version of the formula (as in
\cref{example:dr}) can be handled efficiently. For instance,
\cref{example:first} cannot be handled by \textsc{ForcLift} because there is no
root binding class, i.e., the two root variables belong to different equivalence
classes with respect to the binding relationship. The clauses in this expanded
formula are then partitioned into three parts based on whether the
transformation introduced constants, constraints, or both. The conditions ensure
that these parts can be treated independently.

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap E = \emptyset$ ensures that no pairs of variables in $W$ are
  constrained to be distinct since that would result in an $x \ne x$ constraint
  after substitution. Similarly, we want to avoid variables in $W$ that have
  inequality constraints with constants: after the substitution, such
  constraints transform into inequality constraints between two constants. In
  this case, both subsets of $V$ satisfy the conditions, and
  \cref{line:generation} generates two clauses:
  \begin{multline*}
    (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg \predicate(x, Y), \neg \predicate(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

@@

\Cref{thm:correctness1} is equivalent to Proposition~3 by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11}. For the proofs of this and other
theorems, see supplementary material.

@@

\subsubsection{Caching}
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} uses a cache to check if a
formula is identical to one of the formulas that have already been fully
compiled. To facilitate recursion, we extend the caching scheme to include
formulas encountered before but not fully compiled yet. Formally, we define a
\emph{cache} as a map from integers (e.g., hash codes) to sets of pairs of the
form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG node.

@@

We compare \textsc{Crane} and
\textsc{ForcLift}\footnote{\url{https://dtaid.cs.kuleuven.be/wfomc}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
on their ability to count various classes of functions. We chose this class of
instances because of its simplicity and the inability of publicly available
WFOMC algorithms to solve many such counting problems. Note that, except for a
particular version of \textsc{FastWFOMC}, other exact WFOMC algorithms cannot
solve any instances \textsc{ForcLift} fails on. First, we describe how to
express these function-counting problems in first-order logic. \textsc{ForcLift}
then translates these sentences in first-order logic to formulas as in
\cref{def:formula}.

@@

% the process: both for running algorithms and for determining complexities
We consider all sixteen combinations of these properties (injectivity,
surjectivity, partiality, and endo-), omitting duplicate descriptions of the
same function/sequence. We run each algorithm once on each instance.
\textsc{Crane} is run in hybrid search mode until either it finds five solutions
or the search tree reaches height six. \textsc{ForcLift} is always run until it
terminates. If successful, \textsc{Crane} returns one or more sets of function
definitions, and \textsc{ForcLift} returns a circuit, which can similarly be
interpreted as a function definition. We then assess the complexity of each
solution by hand and pick the best in case \textsc{Crane} returns several
solutions of varying complexities.

@@

Recall that, in the case of \textsc{ForcLift}, solving a (W)FOMC problem
consists of two parts: compilation (via search) and propagation. (A similar
dichotomy applies to \textsc{Crane} as well.) The complexity of the former
depends only on the formula and so far has received very little attention except
for the work by Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15}. The
complexity of the latter, on the other hand, is a function of domain sizes and
can be determined without measuring runtime. We establish the asymptotic
complexity of a solution by counting the number of arithmetic operations needed
to follow the definitions of constituent functions without recomputing the same
quantity multiple times. In particular, we assume that each function call and
binomial coefficient is computed at most once, and computing $\binom{n}{k}$
takes $\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on each $f(i, j)$.

Let $m = |\Gamma|$ and $n = |\Delta|$ be domain sizes. We summarise the results
in \cref{tbl:results}, where we compare the solutions found by both algorithms
to the manually-constructed ways of computing the same quantities that have been
proposed before.\footnote{\url{https://oeis.org},
  \url{https://scicomp.stackexchange.com/q/30049}} Examining the last two
columns of the table, we see that the two algorithms perform equally well on
instances that could already be solved by \textsc{ForcLift}. However,
\textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See supplementary material for
the exact solutions produced by \textsc{Crane}.

@@

\begin{figure}[t]
  \includegraphics{plot}
  \caption{The runtime of \textsc{FastWFOMC} when counting permutations,
    together with a quartic regression model fitted on all of the data and a
    linear model fitted on the first half of the data
    points}\label{fig:fastwfomc}
\end{figure}

@@

\section{Conclusion and Future Work}\label{sec:conclusion}
\setlength{\multlinegap}{0pt}

This paper presents \textsc{Crane}---an extension of
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} that benefits from a more
general version of domain recursion and support for graphs with cycles. Our
results in \cref{sec:results} reveal a range of counting problems that became
newly feasible as a result of these enhancements, including instances outside of
currently-known domain-liftable fragments such as \Ctwo{}. (Although we focused
on unweighted counting, \textsc{ForcLift}'s support for weights trivially
transfers to \textsc{Crane} too.) The common thread across these newly liftable
problems is a version of partial injectivity. Thus, we formulate the following
conjecture.
\begin{conjecture}
  Let \IFO{} be the class of formulas in first-order logic that contain clauses
  with at most two variables as well as any number of copies of
  \begin{align*}
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{i}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{i-1}, Y, X_{i+1}, \dots, X_{n}) \Rightarrow X_{i} = Y) \land{}
    \end{multlined}\\
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{j}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{j-1}, Y, X_{j+1}, \dots, X_{n}) \Rightarrow X_{j} = Y)
    \end{multlined}
  \end{align*}
  for some predicate $\predicate/n$, domains
  $\Delta_{1}, \dots, \Delta_{n}$, and $i, j \in \{\, 1, \dots, n \,\}$. Then
  \IFO{} is liftable by \textsc{Crane}.
\end{conjecture}
Since most of the instances in \cref{sec:results} are in \Ctwo{}, we can also
conjecture that \Ctwo{} is liftable by \textsc{Crane}.

The most important direction for future work is to automate the process in
\cref{fig:processcrane}. First, we need a way to find the base cases for the
recursive definitions produced by \textsc{Crane}. Second, since the first
solution found by \textsc{Crane} is not always optimal in terms of its
complexity, an automated way to determine the asymptotic complexity of a
solution would be helpful as well. Third, as we saw at the end of
\cref{example:interpretation}, the functions constructed by \textsc{Crane} can
often benefit from elementary algebraic simplifications. This process can be
automated using a computer algebra system. Furthermore, note that having the
solution to a (W)FOMC problem expressed in terms of functions enables many new
possibilities such as
\begin{enumerate*}[label=(\roman*)]
  \item the use of more sophisticated simplification techniques,
  \item asymptotic analysis of, e.g., how the model count grows as the domain
  size goes to infinity, and
  \item answering questions parameterised by domain sizes, e.g., `how big does
  domain $\Delta$ have to be for the probability of event $E$ to be above
  \SI{95}{\percent}?'
\end{enumerate*}
In addition to the potential impact on areas of artificial intelligence (AI)
like statistical relational AI~\cite{DBLP:series/synthesis/2016Raedt},
\textsc{Crane} could be beneficial to research in combinatorics as
well~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

>>>>

@@

In this work, we show how to relax these restrictions in a way that results in
an algorithm capable of handling more instances in a lifted manner. The ideas
presented in this paper are implemented in
\textsc{Crane}\footnote{\url{https://github.com/dilkas/crane}}\textsuperscript{,}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}---an
extension of the (W)FOMC algorithm \textsc{ForcLift}. The differences in how
these algorithms operate are depicted in \cref{fig:process}. Compilation is
performed by applying various \emph{(compilation) rules} to the input (or some
derivative) formula, gradually constructing a circuit (in the case of
\textsc{ForcLift}) or a graph (in the case of \textsc{Crane}). \textsc{ForcLift}
applies compilation rules via greedy search, whereas \textsc{Crane} also
supports a hybrid search algorithm that applies some rules greedily and some
using breadth-first search.\footnote{In the current implementation, the rules
  applied non-greedily are: atom counting, inclusion-exclusion, independent
  partial groundings, Shannon decomposition, shattering, and two new rules
  described in \cref{sec:dr,sec:ref}. See previous work for more information
  about the rules~\cite{DBLP:conf/ijcai/BroeckTMDR11} and the search
  algorithm~\cite{dilkas2023generalising}.} This alternative was introduced
because there is no reason to expect greedy search to be complete. Another
difference is that---in \textsc{Crane}---the product of compilation is not
directly evaluated but transformed into a collection of functions on domain
sizes. Hence, our approach is reminiscent of previous work on lifted inference
via compilation to C++ programs~\cite{DBLP:conf/kr/KazemiP16} and the broader
area of functional
synthesis~\cite{DBLP:conf/cav/GoliaRM20,DBLP:conf/pldi/KuncakMPS10}.

@@

Our representation of FOMC instances builds on the format used by
\textsc{ForcLift}, some aspects of which are described by Van den Broeck et
al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate
sentences in a variant of function-free many-sorted first-order logic with
equality to this internal format. We use lowercase Latin letters for predicates
(e.g., $\predicate$) and constants (e.g., $x$), uppercase Latin letters for
variables (e.g., $X$), and uppercase Greek letters for domains (e.g., $\Delta$).
Sometimes we write predicate $\predicate$ as $\predicate/n$, where
$n \in \mathbb{N}^{+}$ is the \emph{arity} of $\predicate$. An \emph{atom} is
$\predicate(x_1, \dots, x_n)$ for some predicate $\predicate/n$ and terms
$x_{1}, \dots, x_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation thereof (denoted by
$\neg \predicate(x_1, \dots, x_n)$). Let $\mathcal{D}$ be the set of all
relevant domains. Initially, $\mathcal{D}$ contains all domains mentioned by the
input formula. During compilation, new domains are added to $\mathcal{D}$. Each
new domain is interpreted as a subset of another domain in $\mathcal{D}$.

@@

Let $\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain size
  function} that maps each domain to a non-negative integer.

% By a relation between two (potentially distinct) sets, we mean any
% subset of their Cartesian product. Thus, if both sets have cardinality two,
% their Cartesian product has cardinality four, and this product set has 2^4 =
% 16 subsets. To check whether a subset S is a model, we interpret S as the set
% of all pairs (X, Y) for which p(X, Y) is true. Each of the two clauses in the
% example (c_1 and c_2) imposes certain constraints on S. For example, for
% domains Gamma = Delta = {1, 2}, clause c_1 requires that:
% 1) either (1, 1) or (1, 2) is not in S,
% 2) either (2, 1) or (2, 2) is not in S.

\begin{example}
  Let $\phi$ be as in \cref{example:first} and $\Gamma = \Delta = \{\,1, 2\,\}$,
  i.e., $\sigma(\Gamma) = \sigma(\Delta) = 2$. There are $2^{2 \times 2} = 16$
  possible relations between $\Gamma$ and $\Delta$, i.e., subsets of
  $\Gamma \times \Delta$. Let us count how many of them satisfy the conditions
  imposed on predicate $\predicate$. The empty relation does. All four relations
  of cardinality one (e.g., $\{\, (1, 1) \,\}$) do too. Finally, there are two
  relations of cardinality two---$\{\, (1, 1), (2, 2) \,\}$ and
  $\{\, (1, 2), (2, 1) \,\}$---that satisfy the conditions as well. Therefore,
  the FOMC of $(\phi, \sigma)$ is 7. Incidentally, the FOMC of $\phi$ counts
  partial injections from $\Gamma$ to $\Delta$. We will continue to use the
  problem of counting partial injections as the main running example.
\end{example}

@@

% Original domain recursion
Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} imposes stringent conditions
on the input formula to ensure that its expanded version (as in
\cref{example:dr}) can be handled efficiently. \Cref{example:first} cannot be
handled by \textsc{ForcLift} because there is no root binding class, i.e., the
two root variables belong to different equivalence classes with respect to the
binding relationship.

% The clauses in this expanded
% formula are then partitioned into three parts based on whether the
% transformation introduced constants, constraints, or both. The conditions ensure
% that these parts can be treated independently.

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap E = \emptyset$ ensures that no pairs of variables in $W$ are
  constrained to be distinct since that would result in an $x \ne x$ constraint
  after substitution. Similarly, we want to avoid variables in $W$ that have
  inequality constraints with constants. In this case, both subsets of $V$
  satisfy the conditions, and \cref{line:generation} generates two clauses:
  \begin{multline*}
    (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg \predicate(x, Y), \neg \predicate(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

@@

\Cref{thm:correctness1} is equivalent to Proposition~3 by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11}. For the proofs of this and other
theorems, see \cref{sec:proofs}.

@@

\subsubsection{Caching}
\textsc{ForcLift} uses a cache to check if a formula is identical to one of the
formulas that have already been fully compiled. To facilitate recursion, we
extend the caching scheme to include formulas encountered before but not fully
compiled yet. Formally, we define a \emph{cache} as a map from integers (e.g.,
hash codes) to sets of pairs of the form $(\phi, v)$, where $\phi$ is a formula,
and $v$ is an FCG node.

@@

We compare \textsc{Crane} and
\textsc{ForcLift}\footnote{\url{https://dtaid.cs.kuleuven.be/wfomc}} on their
ability to count various classes of functions. We chose this class of instances
because of its simplicity and the inability of publicly available WFOMC
algorithms to solve many such counting problems. Note that, except for a
particular version of \textsc{FastWFOMC}, other exact WFOMC algorithms cannot
solve any instances \textsc{ForcLift} fails on. First, we describe how to
express these function-counting problems in first-order logic. \textsc{ForcLift}
then translates these sentences in first-order logic to formulas as in
\cref{def:formula}.

@@

% the process: both for running algorithms and for determining complexities
We consider all sixteen combinations of these properties (injectivity,
surjectivity, partiality, and endo-), omitting duplicate descriptions of the
same function, e.g., the number of partial $\Gamma \to \Delta$ bijections is the
same as the number of $\Delta \to \Gamma$ injections. We run each algorithm once
on each instance. \textsc{Crane} is run in hybrid search mode until either it
finds five solutions or the search tree reaches height six. \textsc{ForcLift} is
always run until it terminates. If successful, \textsc{Crane} returns one or
more sets of function definitions, and \textsc{ForcLift} returns a circuit,
which can similarly be interpreted as a function definition. We then assess the
complexity of each solution by hand and pick the best in case \textsc{Crane}
returns several solutions of varying complexities. In our experiments,
\textsc{Crane} found at most four solutions per problem instance, and most
solution had the same complexity.

% Automatically determining the complexity of a solution generated by
% Crane is another area for future work. Alternatively, one could evaluate
% several solutions in parallel until one of them produces the model count. In
% our experiments, Crane found at most four solutions per problem instance. Most
% had the same complexity, although some were less efficient (e.g., cubic
% instead of linear or quintic instead of cubic).

@@

Recall that, in the case of \textsc{ForcLift}, solving a (W)FOMC problem
consists of two parts: compilation (via search) and propagation. (A similar
dichotomy applies to \textsc{Crane} as well.) The complexity of the former
depends only on the formula and so far has received very little attention except
for the work by Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15}. The
complexity of the latter, on the other hand, is a function of domain sizes and
can be determined without measuring runtime. We establish the asymptotic
complexity of a solution by counting the number of arithmetic operations needed
to follow the definitions of constituent functions without recomputing the same
quantity multiple times.\footnote{Although this is not done by default, one
  could also use arbitrary-precision arithmetic and consider bit complexity
  instead.} In particular, we assume that each function call and binomial
coefficient is computed at most once, and computing $\binom{n}{k}$ takes
$\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on each $f(i, j)$.

Let $m = |\Gamma|$ and $n = |\Delta|$ be domain sizes. We summarise the results
in \cref{tbl:results}, where we compare the solutions found by both algorithms
to the manually-constructed ways of computing the same quantities that have been
proposed before.\footnote{\url{https://oeis.org},
  \url{https://scicomp.stackexchange.com/q/30049}} Examining the last two
columns of the table, we see that the two algorithms perform equally well on
instances that could already be solved by \textsc{ForcLift}. However,
\textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See \cref{sec:solutions} for
the exact solutions produced by \textsc{Crane}.

@@

\begin{figure}[t]
  \includegraphics{plot}
  \caption{The runtime of \textsc{FastWFOMC} when counting permutations
    (indicated by points), together with a quartic regression model fitted on
    all of the data and a linear model fitted on the first half of the data
    points}\label{fig:fastwfomc}
\end{figure}

@@

\section{Discussion}\label{sec:conclusion}
\setlength{\multlinegap}{0pt}

This paper presents \textsc{Crane}---an extension of
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} that benefits from a more
general version of domain recursion and support for graphs with cycles. In
\cref{sec:results}, we listed a range of counting problems that became newly
feasible as a result, including instances outside of currently-known
domain-liftable fragments such as \Ctwo{}. (Although we focused on unweighted
counting, \textsc{ForcLift}'s support for weights trivially transfers to
\textsc{Crane} too.) The common thread across these newly liftable problems is a
version of partial injectivity. Thus, we formulate the following conjecture for
a class of formulas with an injectivity-like constraint on at most two
parameters of a predicate $\predicate$.

\begin{conjecture}\label{conj}
  Let \IFO{} be the class of formulas in first-order logic that contain clauses
  with at most two variables as well as any number of copies of
  \begin{align*}
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{i}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{i-1}, Y, X_{i+1}, \dots, X_{n}) \Rightarrow X_{i} = Y) \land{}
    \end{multlined}\\
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{j}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{j-1}, Y, X_{j+1}, \dots, X_{n}) \Rightarrow X_{j} = Y)
    \end{multlined}
  \end{align*}
  for some predicate $\predicate/n$, domains
  $\Delta_{1}, \dots, \Delta_{n}$, and $i, j \in \{\, 1, \dots, n \,\}$. Then
  \IFO{} is liftable by \textsc{Crane}.
\end{conjecture}

Note that the unsolved instance in \cref{tbl:results} does not invalidate
\cref{conj} for two reasons. First, in our experiments, the depth of the search
procedure is limited, so there is a possibility that increasing the depth would
make the algorithm succeed. Second, the conjecture assumes distinct domains,
while the unsolved instance pertains to endofunctions.

% In general, our results demonstrate that Crane
% performs better when all domains are distinct. A solution for distinct domains
% can always be applied to the case when some domains are equal.

% ===========================================

% Since most of the instances in \cref{sec:results} are in \Ctwo{}, we can also
% conjecture that \Ctwo{} is liftable by \textsc{Crane}.

% Every sentence in C2 can be rewritten to avoid counting quantifiers by
% introducing more variables. For example, to claim that there are exactly two
% Xs that satisfy p(X), one would introduce two existentially-quantified
% variables X_1 and X_2 such that X_1 != X_2, p(X_1), and p(X_2) and add a
% constraint that any other variable X_3 such that p(X_3) has to be equal to
% either X_1 or X_2.

The most important direction for future work is to automate the process in
\cref{fig:processcrane}. First, we need a way to find the base cases for the
recursive definitions produced by \textsc{Crane}. Second, since the first
solution found by \textsc{Crane} is not always optimal in terms of its
complexity, an automated way to determine the asymptotic complexity of a
solution would be helpful as well. Third, as we saw at the end of
\cref{example:interpretation}, the functions constructed by \textsc{Crane} can
often benefit from elementary algebraic simplifications. This process can be
automated using a computer algebra system. Furthermore, note that having the
solution to a (W)FOMC problem expressed in terms of functions enables many new
possibilities such as
\begin{enumerate*}[label=(\roman*)]
  \item the use of more sophisticated simplification techniques,
  \item asymptotic analysis of, e.g., how the model count grows as the domain
  size goes to infinity, and
  \item answering questions parameterised by domain sizes, e.g., `how big does
  domain $\Delta$ have to be for the probability of event $E$ to be above
  \SI{95}{\percent}?'
\end{enumerate*}

\section*{Acknowledgments}

We thank Anna L.D. Latour and Kuldeep S. Meel for comments on earlier drafts of
this paper. Most of the work for this paper was done while the first author was
a PhD student at the University of Edinburgh. The first author was supported by
the EPSRC Centre for Doctoral Training in Robotics and Autonomous Systems,
funded by the UK Engineering and Physical Sciences Research Council (grant
EP/L016834/1). The second author was partly supported by a Royal Society
University Research Fellowship, UK, and partly supported by a Grant from the
UKRI Strategic Priorities Fund, UK to the UKRI Research Node on Trustworthy
Autonomous Systems Governance and Regulation (EP/ V026607/1, 2020–2024). For the
purpose of open access, the authors have applied a Creative Commons Attribution
(CC BY) licence to any Author Accepted Manuscript version arising from this
submission.

\appendix
\section{Proofs}\label{sec:proofs}

We begin with a few definitions that formalise the notion of a model. For any
set $S$, let $2^{S}$ denote its power set and
$S^{\ast} \coloneqq \bigcup_{i=0}^{\infty} S^{i}$ the (infinite) set of
tuples---of any finite length---of elements of $S$. For example,
${\{\,a, b\,\}}^{\ast} = \{\, (), (a), (b), (a, a), \dots \,\}$. For any
$n \in \mathbb{N}_{0}$, let $[n] \coloneqq \{\, 1, \dots, n \,\}$, e.g.,
$[0] = \emptyset$, and $[2] = \{\, 1, 2 \,\}$. Extending the notation for
substitution introduced in \cref{sec:recprelims}, for a formula $\phi$, constant
$c$, and set of variables $V$, let $\phi[c/V]$ denote $\phi$ with all
occurrences of all variables in $V$ replaced with $c$. Let $\Preds$ be the
function that maps any clause or formula to the set of predicates used in it.
With each formula $\phi$, we associate a map
$\pi_{\phi}\colon \Preds(\phi) \to {\Doms(\phi)}^{\ast}$ s.t., for each
predicate $\predicate/n \in \Preds(\phi)$, we have that
$\pi_{\phi}(\predicate) \in {\Doms(\phi)}^{n}$. This map makes explicit the idea
that each $n$-ary predicate is associated with an $n$-tuple of domains. In the
case of formula $\phi$ from \cref{example:first},
$\pi_{\phi} = \{\, \predicate \mapsto (\Gamma, \Delta) \,\}$.

\begin{definition}\label{def:model}
  Let $\phi$ be a formula and $\sigma$ a domain size function. A \emph{model} of
  $(\phi, \sigma)$ is a map
  $\mathfrak{M}\colon \Preds(\phi) \to 2^{\mathbb{N}_{0}^{\ast}}$ s.t.\ the
  following two conditions are satisfied.
  \begin{enumerate}
    \item For all predicates $\predicate/n \in \Preds(\phi)$ with
          $\pi_{\phi}(\predicate) \coloneqq {(\Delta_{i})}_{i=1}^{n}$ for some
          domains $\Delta_{i} \in \Doms(\phi)$,
    \begin{equation}\label{eq:subset}
      \mathfrak{M}(\predicate) \subseteq \prod_{i=1}^{n}[\sigma(\Delta_{i})].\footnote{For simplicity, \cref{def:model} ignores constants. To include constants, one would replace $[\sigma(\Delta_{i})]$ with a set that contains all constants associated with domain $\Delta_{i}$, extended with enough new elements to make its cardinality $\sigma(\Delta_{i})$.}
    \end{equation}
    \item $\mathfrak{M} \models \phi$ according to the (non-exhaustive)
          recursive definition below. Let $\chi$ and $\psi$ be formulas,
          $\Delta$ a domain, $c \in \Delta$ a constant, $\predicate/n$ a
          predicate associated with domains $\Delta_{1},\dots,\Delta_{n}$, and
          $t \in \prod_{i=1}^{n} \Delta_{i}$ a tuple of constants.
    \begin{itemize}
      \item $\mathfrak{M} \models \predicate(t)$ if and only if (iff)
            $t \in \mathfrak{M}(\predicate)$.
      \item $\mathfrak{M} \models \neg\chi$ iff $\mathfrak{M} \not\models \chi$.
      \item $\mathfrak{M} \models \chi \land \psi$ iff
            $\mathfrak{M} \models \chi$ and $\mathfrak{M} \models \psi$.
      \item $\mathfrak{M} \models \chi \lor \psi$ iff
            $\mathfrak{M} \models \chi$ or $\mathfrak{M} \models \psi$.
      \item $\mathfrak{M} \models (\forall X \in \Delta\text{. }\chi)$ iff
            $\mathfrak{M} \models \chi[x/\{\, X \,\}]$ for all constants
            $x \in \Delta$.
      \item $\mathfrak{M} \models (\forall X \in \Delta\text{.
            } X \ne c \Rightarrow \chi)$ iff
            $\mathfrak{M} \models \chi[x/\{\, X \,\}]$ for all
            $x \in \Delta \setminus \{\, c \,\}$.
      \item $\mathfrak{M} \models (\forall X,Y \in \Delta\text{.
            } X \ne Y \Rightarrow \chi)$ iff
            $\mathfrak{M} \models \chi[x/\{\, X \,\}][y/\{\, Y \,\}]$ for all
            $x,y \in \Delta$ s.t. $x \ne y$.
    \end{itemize}
  \end{enumerate}
\end{definition}

\begin{theorem}[Correctness of $\GDR$]
  Let $\phi$ be the formula used as input to \cref{alg:domainrecursion},
  $\Omega \in \mathcal{D}$ the domain selected on \cref{line:condition}, and
  $\phi'$ the formula constructed by the algorithm for $\Omega$. Suppose that
  $\Omega \ne \emptyset$. Then $\phi \equiv \phi'$.
\end{theorem}
\begin{proof}
  Let $x$ be the constant introduced on \cref{line:constant} and $c$ a clause of
  $\phi$ selected on \cref{line:forclause} of the algorithm. We will show that
  $c$ is equivalent to the conjunction of clauses added to $\phi'$ on
  \cref{line:conditions,line:generation}.

  If $c$ has no variables with domain $\Omega$,
  \cref{line:conditions,line:generation} add a copy of $c$ to $\phi'$. If $c$
  has one variable with domain $\Omega$, say, $X$, then
  $c \equiv \forall X \in \Omega\text{. } \psi$ for some formula $\psi$ with one
  free variable $X$. In this case, \cref{line:conditions,line:generation} create
  two clauses: $\psi[x/\{\, X \,\}]$ and $\forall X \in \Omega\text{.
  } X \ne x \Rightarrow \psi$. It is easy to see that their conjunction is
  equivalent to $c$ provided that $\Omega \ne \emptyset$ and so $x$ is a
  well-defined constant.
